Creating Named Valleys Dataset

### 2023-06-16 scripts/Barneby_Lives.Rmd
bringing in dististaze from sos-gb

While in theory all of the tasks in this script could be automated, doing so would be an unconscionable action against local taxonomic authorities. While several database options for updating collections (and spell checking nomen) which are current and well curated (e.g. Kew Plants of the World) exist, we feel many proposals and combinations require interpretation by those in the regions attempting to understand the material at hand.

@@

```{r Libraries, message=FALSE, warning=FALSE}
#install.packages("WorldFlora", "elevatr")
library(tidyverse)
library(sf)
#library(parzer)
library(tigris)
#library(WorldFlora)
library(elevatr)
```

@@

@@

>>>>

While in theory all of the tasks in this script could be automated, doing so would be an unconscionable action against local taxonomic authorities. While several database options for updating collections (and spell checking names) which are current and well curated (e.g. Kew Plants of the World) exist, we feel many proposals and combinations require interpretation by those in the regions attempting to understand the material at hand.

@@

```{r Libraries, message=FALSE, warning=FALSE}
#install.packages("WorldFlora", "elevatr")
library(tidyverse)
library(sf)
#library(parzer)
library(tigris)
#library(WorldFlora)
library(elevatr)
library(googlesheets4)
```

@@

An alternative format for loading data is through Google Sheets. This is what we use within our group.

```{r load through google sheets}

library(googlesheets4)

# read in data from the sheet to process
input <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
                    sheet = 'Data Entry - Examples') %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

# determine whether these data have already been processed by the script, using
# a unique combination of collection name and collection code. 
processed <- read_sheet('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M',
                        sheet = 'Processed - Examples') %>% 
  select(Collection_number, Primary_Collector) %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number))

df <- filter(input, ! UNIQUEID %in% processed$UNIQUEID ) %>% 
  select(-UNIQUEID)

rm(processed, input)

# we will add these data onto our final sheet. 
sheet_append('1iOQBNeGqRJ3yhA-Sujas3xZ2Aw5rFkktUKv3N_e4o8M', 
             sheet = 'Processed - Examples', data = df) 
```

@@

**distance from named place**

```{r}
sites <- data.frame(
  longitude_dd = runif(15, min = -120, max = -100), 
  latitude_dd = runif(15, min = 35, max = 48)
  ) %>% 
  st_as_sf(coords = c('longitude_dd', 'latitude_dd'))

places <- st_read('../geodata/places/places.shp')

head(places)
```

```{r}

#' gather distance and azimuth from shop to center of town
#'
#' Help provide some simple context between the building and the 
#' @param x an sf/tibble/dataframe of locations with associated nearest locality data
#' @param places_data an sf/dataframe/dataframe which contains coordinates for the locations in the data
distAZE <- function(x, places_data){

  locality <- sf::st_drop_geometry(x)
  locality <- locality[1, 'Locality']

  focal <- places_data[grep(locality, places_data$ID), ]
  if(nrow(focal) > 1){
    union_loc <- sf::st_union(x) |> sf::st_point_on_surface()
    focal <- focal[sf::st_nearest_feature(union_loc, focal),]
  }

  location_from <- sf::st_centroid(focal)

  location_from <- sf::st_transform(location_from, 5070)
  x_planar <- sf::st_transform(x, 5070)
  distances <- sf::st_distance(location_from, x_planar, which = 'Euclidean')

  azy <- nngeo::st_azimuth(
    location_from, 
    x_planar
  )

  distances <- data.frame(
    st_drop_geometry(x),
    Distance = round(as.numeric(distances / 1609.34), 1),
    Azimuth = round(as.numeric(azy), 0),
    geometry = x[,'geometry']
  )

}

### 2023-06-16 scripts/valleys_create.Rmd
bringing in dististaze from sos-gb

The 'Mountains Inventory version 2' dataset generated by Global Mountain Biodiversity Assessment contains a standardized naming schema for all of the worlds mountains. We are interested in developing a similar named schema for many of the valleys of Western North America.

In order to identify drainage areas, 
USGS Watershed Boundary Dataset (WBD) was used as a source of all major watersheds [The National Map](https://apps.nationalmap.gov/downloader/#/).

@@

```{r mountains}
mtns <- st_read('../geodata/mountains/mountains.shp', quiet = T)
mtn_bbox <- mtns %>% 
  st_transform(4269) %>% 
  st_bbox()

mtns <- st_make_valid(mtns)
```

HU 10 dataset was used as the source for watersheds wherein we would name valleys. It was cropped to the extent of the study area.

@@

hu2 <- st_read(dsn = pbasin, layer = 'WBDHU2')  %>% 
  select(HU2_NAME = name) %>% 
  st_crop(., mtn_bbox)

hu10 <- st_read(dsn = pbasin, layer = 'WBDHU10')  %>% 
  select(HU10_NAME = name) %>% 
  st_crop(., mtn_bbox)

hu10 <- hu10[st_intersects(hu10, st_union(hu2)) %>% lengths > 0, ]

hu10 <- st_transform(hu10, 5070) %>% 
  st_simplify(dTolerance = 75) %>% 
  st_transform(4326) # around 1/9 the size of the original

hu10 <- st_make_valid(hu10)
```

```{r identify watersheds in the great basin area}

gb <- filter(hu2, HU2_NAME == 'Great Basin Region') %>% 
  st_transform(4326)

gb10 <- hu10[st_intersects(hu10, st_union(gb)) %>% lengths > 0, ]
```

```{r overlay with gb mtns}

mtns_gb <- mtns[st_intersects(mtns, st_union(gb10)) %>% lengths > 0, ]
gb10_er <- rmapshaper::ms_erase(gb10, mtns)

ggplot() + 
  geom_sf(data = gb10_er) +
  geom_sf(data = mtns_gb, fill = 'purple', alpha = 0.3)
```

```{r identify western watersheds}
west_pl <- c('California Region', 'Pacific Northwest Region', 
             'Lower Colorado Region', 'Great Basin Region', 'Upper Colorado Region', 
             'Rio Grande Region')
west <- filter(hu2, HU2_NAME %in% west_pl) %>% 
  st_transform(4326)

west10 <- hu10[st_intersects(hu10, st_union(west)) %>% lengths > 0, ]
```

```{r overlay with western mtns}

mtns_west <- mtns[st_intersects(mtns, st_union(west)) %>% lengths > 0, ]
west10 <- st_crop(west10, mtns_west)
west10_er <- rmapshaper::ms_erase(west10, mtns_west)

ggplot() + 
  geom_sf(data = west10) +
  geom_sf(data = mtns_west, fill = 'purple', alpha = 0.3)
```

**First two authors in alphabetical order**

>>>>

The 'Mountains Inventory version 2' dataset generated by Global Mountain Biodiversity Assessment contains a standardized naming schema for all of the worlds mountains. We are interested in developing a simpler named schema for many of the valleys and watersheds of Western North America.

USGS Watershed Boundary Dataset (WBD) was used to identify major watersheds [The National Map](https://apps.nationalmap.gov/downloader/#/).

@@

```{r mountains}
mtns <- st_read('../geodata/mountains/mountains.shp', quiet = T)
mtn_bbox <- mtns %>% 
  st_transform(4269) %>% 
  st_bbox()
```

@@

hu2 <- st_read(dsn = pbasin, layer = 'WBDHU2', quiet = T)  %>% 
  select(HU2_NAME = name) %>% 
  st_crop(., mtn_bbox)

hu8 <- st_read(dsn = pbasin, layer = 'WBDHU8', quiet = T)  %>% 
  select(HU8_NAME = name) %>% 
  st_crop(., mtn_bbox)

hu8 <- st_join(hu2, hu8) 
```

```{r}

ggplot() + 
  geom_sf(data = hu8)

gb <- filter(hu2, name == 'Great Basin Region')
```
